Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations: the Oziris algorithm

### 2016-11-04 ParisEM.tex
Small typos
Major changes in the algorithm. Changes in notations with zeta as the random variable for bridges. Attempt of harmonizing notations of indexes k, i, l. Major modifications in the proof of lemma 1. Putting proof of lemma 2 in appendix with minor modifications (adding a step that was not straightforward for me). Addition of intermetiate algorithm between the one of Paris and ours

\newcommand{\x}[2]{x_{#1}^{(#2)}}
\newcommand{\w}[2]{w_{#1}^{(#2)}}
\newcommand{\tw}[2]{\tilde{w}_{#1}^{(#2)}}
\newcommand{\p}[2]{\xi_{#1}^{(#2)}}
\newcommand{\tp}[2]{\tilde{\xi}_{#1}^{(#2)}}
\newcommand{\ta}[2]{\tau_{#1}^{(#2)}}
\newcommand{\rmd}{\mathrm{d}}
\newcommand{\eqsp}{\;}
\newcommand{\1}{\mathrm{1}}
\newcommand{\com}[1]{{\color{gray} // #1}}
\newcommand{\mP}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\qk}{q_{k}}
%\newcommand{\qk}{q^{\Delta t_k}_{\theta}}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}

@@

For all $1\le k \le n$, the distribution of $X_{k} $ conditional on $X_{k-1}$ has a density $\qk(X_{k-1},\cdot) :=q(\theta,\Delta_k,X_{k-1},\cdot),~\Delta_k:=t_k-t_{k-1}$ with respect to a reference measure $\mu$ on $\mathbb{R}^d$.\\

@@

p_{\theta}(x_{0:n},Y_{0:n}) = \chi(x_0)g_{0}(x_0)\prod^{n}_{k=1}\qk(x_{k-1},x_k)g_{k}(x_k)\eqsp.

@@

+ \sum_{k=1}^n\mathbb{E}_{\theta_p}\left[\log \left(\qk(X_{k-1},X_k)g_k(X_k)\right)\middle|Y_{0:n}\right] \eqsp.

@@

H_{n} =  \sum_{k=1}^n\mathbb{E}\left[h_k(X_{k-1},X_k)\middle|Y_{0:n}\right]\eqsp,

@@

when the transition density of the hidden states is not available explicitly and where $\{h_k\}_{k=1}^n$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$.

@@

The PaRIS algorithm uses the same decomposition as the FFBS algorithm introduced in \cite{doucetgodsillandrieu:2000} and the FFBSI algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions. 

@@

The algorithm is based on sufficient statistics $\tau^i_k$, for $1\le i\le N$ and $0\le k \le n$ starting with $\tau^i_0 = 0$ for all $1\le i\le N$. Let $\tilde{N}\ge 1$. Then, at each time step $1\le k \le n$ these statistics are update according to the following steps.

@@

\item \label{it:PaRIS:sampleindex} For all $1\le i \le N$, sample independently $J_k^{i,\ell}$ in $\{1,\ldots,N\}$ for $1\le \ell \le \widetilde N$ with probabilities $\Lambda_{k-1}^N(i,\cdot)$, given by

@@

\Lambda_{k-1}^N(i,\ell) = \frac{\omega^{\ell}_{k-1} \qk(\xi^{\ell}_{k-1},\xi_k^{i})}{\sum_{\ell=1}^N\omega^{\ell}_{k-1} \qk(\xi^{\ell}_{k-1},\xi_k^{i})}\eqsp,\quad 1\le \ell\le N\eqsp.

@@

\tau^{i}_{k} := \frac{1}{\widetilde{N}} \sum^{\widetilde{N}}_{\ell=1} \left\{ \tau^{J_k^{i,\ell}}_{k-1} + h_{k} \left(\xi^{J_k^{i,\ell}}_{k-1}, \xi^{i}_{k}\right)  \right\}\eqsp.

@@

As proved in \cite{olsson:westerborn:2016}, the algorithm is asymptotically consistent (as $N$ goes to infinity) for any fixed precision parameter $\tilde N$. However, there is a significant qualitative difference between the cases $\tilde{N} = 1$ and $\tilde{N} \geq 2$.

As for the FFBSi algorithm,  when there exists $0<q^{\Delta t_k}<\sigma_+$, the PaRIS algorithm may be implemented with $\mathcal{O}(N)$ complexity using the procedure of \cite{douc:garivier:moulines:olsson:2011}.
As $q^{\Delta t_k}(\xi^{\ell}_{k-1},\xi_k^{i})$ is not available explicitly for all $1\le i,\ell\le N$, the next section introduces GPE to replace it by $\widehat{q}^{\Delta t_k}(\xi^{\ell}_{k-1},\xi_k^{i};\Lambda)$, where $\Lambda$ is a random variable such that
\[
\mathbb{E}\left[\widehat{q}^{\Delta t_k}(\xi^{\ell}_{k-1},\xi_k^{i};\Lambda)\middle| \mathcal{F}_k^N\right] = q^{\Delta t_k}(\xi^{\ell}_{k-1},\xi_k^{i})\eqsp,
\]
where 
\[
\mathcal{F}_k^N = \sigma\left\{Y_{0:k};(\xi^{\ell}_{u},\omega^{\ell}_{u},\tau^{\ell}_{u}), 1\le \ell\le N, 1\le u\le k\right\}\eqsp.
\]
In the case where $\widehat{q}^{\Delta t_k}$ may be upper bounded by $\sigma_+$, step~\eqref{it:PaRIS:sampleindex} is replaced by the acceptance-rejection mechanism given in Algorithm~\ref{alg:AR}. It is proved in Lemma~\ref{lem:AR:unbiased} that replacing $\widehat{q}^{\Delta t_k}$ by its unbiased approximation is a valid alternative to step~\eqref{it:PaRIS:sampleindex}.

\begin{algorithm}
\caption{Random weight Accept-reject-based backward sampling  (\cite{douc:garivier:moulines:olsson:2011})}
\begin{algorithmic}
\FORALL{$i \in 1,\dots,\widetilde N$}
\STATE Set $L =\{1,\ldots,N\}$. 
\WHILE{$L\neq\emptyset$}
\STATE $p = \#L$.
\STATE Sample $I = (I_1,\ldots,I_p)$ independently in $\{1,\ldots,N\}$ with probabilities proportional to $\omega_{k-1}^{\ell}$, $1\le \ell\le N$.
\STATE Sample $U = (U_1,\ldots,U_p)$ independently and independently of $I$ uniformly in $[0,1]$.
\STATE Sample $\Lambda = (\Lambda_1,\ldots,\Lambda_p)$ independently and independently of $I$ and $U$.
\STATE Set $L_p = \emptyset$.
\FORALL{$\ell \in 1,\dots,p$}
\IF{$U_\ell<\widehat{q}(\xi_{k-1}^{I_\ell},\xi_k^{L(\ell)};\Lambda_{\ell})/\sigma_+$}
\STATE $J_k^{L(\ell),i} = I_\ell$.
\ELSE
\STATE $L_p = L_p \cup \{L(\ell)\}$.
\ENDIF
\ENDFOR
\STATE $L = L_p$.
\ENDWHILE
\ENDFOR
\end{algorithmic}
\label{alg:AR}
\end{algorithm}

\begin{lemma}
\label{lem:AR:unbiased}
Let $1\le k\le n$.  For all $1\le i \le N$ and all $1\le j \le \widetilde{N}$, the conditional probability distribution given $\mathcal{F}_k^N$ of the random variable $J_k^{i,j}$ produced by Algorithm~\ref{alg:AR} is  $\Lambda_{k-1}^N(i,\cdot)$. 
\end{lemma}

\section{Ozaki-PaRIS algorithm}
\label{sec:PaRIS:SDE}
In general situations, the PaRIS algorithm cannot be used for stochastic differential equations as $q^{\Delta t_k}_{\theta}$ is unknown. Therefore, the computation of the importance weights $\omega_{k}^{\ell}$ and of the acceptance ratio in Algorithm~\ref{alg:AR} cannot be performed to apply the PaRIS algorithm. Following \cite{olsson:strojby:2011},
an extension of the PaRIS algorithm may be developed by replacing $q^{\Delta t_k}_{\theta}$  by an unbiased estimator. In \cite{olsson:strojby:2011},  GPE are used to estimate $q^{\Delta t_k}_{\theta}$ which relies on a representation of $q^{\Delta t_k}_{\theta}$ as an expectation under the probability measure of a Brownian bridge. In order to improve the Monte Carlo estimation of $q^{\Delta t_k}_{\theta}$,  other diffusion bridges may be used as long as they can be simulated efficiently. A simple idea is to consider a diffusion obtained as an approximation of the target SDE \eqref{eq:target:sde}. \cite{ozaki:1992,shoji:ozaki:1998} proposed a higher order approximation of the drift term $\alpha_{\theta}$ than the Euler scheme. The authors used a linear approximation of the diffusion drift $\alpha_{\theta}$, together with a constant approximation of the volatility over each time interval $(t_{k-1},t_k)$. Following \cite{ozaki:1992,shoji:ozaki:1998}, for all $1\le k\le n$, let $(Y^{k}_t)_{t\ge 0}$ be governed by the following stochastic differential equation for $t\in(t_{k-1},t_k)$:
\begin{equation}
\label{eq:SDE:Ozaki}
Y^k_{t_{k-1}} = X_{k-1}\quad\mbox{and}\quad\rmd Y^k_t = \widetilde{\alpha}^k_{\theta}(Y^k_t)\rmd t + \rmd W_t\eqsp,
\end{equation}
where
\[
\widetilde{\alpha}^k_{\theta}(x) = J_{\alpha_{\theta}}(X_{k-1})x + \alpha_{\theta}(X_{k-1}) - J_{\alpha_{\theta}}(X_{k-1})X_{k-1}
\]
and $J_{\alpha_{\theta}}$ is the Jacobian of $\alpha_{\theta}$. The probability density function of the transition kernel associated with $Y^k$ is written $\widetilde{q}^{\Delta t}_{\theta}$ and is given in Section~\ref{sec:ornstein:bridge}.

Writing $\mathbb{Q}^k_{\theta}$ (resp. $\mathbb{W}^{k}_{\theta}$) the probability measure induced by $(X_s)_{t_{k-1}\le s\le t_k}$  (resp. $(Y_s)_{t_{k-1}\le s\le t_k}$) on $(\mathsf{C},\mathcal{C})$, Girsanov theorem yields:

@@

\frac{\rmd \mathbb{Q}^k_{\theta}}{\rmd \mathbb{W}^{k}_{\theta}}(w) = \mathrm{exp}\left\{H_{\theta}(\omega_{t_k}) - H_{\theta}(\omega_{t_{k-1}}) - \frac{1}{2}\int_{t_{k-1}}^{t_k}\left\{\|\alpha_{\theta}(\omega_s)-\widetilde{\alpha}_{\theta}(\omega_s)\|^2+\Delta H_{\theta}(\omega_s)\right\}\rmd s\right\}\eqsp,

@@

H_{\theta}(x) = A_{\theta}(x) - x'J_{\alpha_{\theta}}(X_{k-1})x/2- (\alpha_{\theta}(X_{k-1}) - J_{\alpha_{\theta}}(X_{k-1})X_{k-1})'x

@@

and $\Delta$ is the Laplace operator. Then, the probability measures conditioned on hitting $X_{k}$ at time $t_k$ satisfy:

@@

\frac{\rmd \mathbb{Q}^{k,X_{k}}_{\theta}}{\rmd \mathbb{W}^{k,X_{k}}_{\theta}}(w) = \frac{\widetilde{q}^{\Delta t_k}_{\theta}(X_{k-1},X_{k})}{q^{\Delta t_k}_{\theta}(X_{k-1},X_{k})}\mathrm{exp}\left\{H_{\theta}(X_k) - H_{\theta}(X_{k-1})\right\} \\
\times \mathrm{exp}\left\{- \frac{1}{2}\int_{t_{k-1}}^{t_k}\left\{\|\alpha_{\theta}(\omega_s)-\widetilde{\alpha}_{\theta}(\omega_s)\|^2+\Delta H_{\theta}(\omega_s)\right\}\rmd s\right\}\eqsp.

@@

\item For all $\theta\in\Theta$ and all $0\le k \le n-1$, $x\mapsto \alpha_{\theta}(x) - \alpha_{\theta}(X_k) - J_{\alpha_{\theta}}(X_{k})(x-X_k)$ is bounded.
\item For all $\theta\in\Theta$ and all $0\le k \le n-1$, $x\mapsto \mathrm{Tr}\left(J_{\alpha_{\theta}}(x) - J_{\alpha_{\theta}}(X_k)\right)$ is bounded, where $\mathrm{Tr}$ is the trace operator.

@@

q^{\Delta t_k}_{\theta}(X_{k-1},X_k) = \widetilde{q}^{\Delta t_k}_{\theta}(X_{k-1},X_k) \eqsp\mathrm{exp}\left\{H_{\theta}(X_k) - H_{\theta}(X_{k-1})-\ell(\theta)\Delta t_k\right\}\\
\times\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta t_k}\phi_{\theta}(\omega_s)\rmd s\right\}\right]\eqsp,

@@

where $\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}$ is the expectation under the law of the bridge process defined by \eqref{eq:SDE:Ozaki}, starting at $X_{k-1}$ at time $0$ and ending at $X_{k}$ at time $\Delta t_k$.
An unbiased estimator of $q^{\Delta t_k}_{\theta}(X_{k-1},X_k)$ is then obtained by estimating $\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta t_k}\phi_{\theta}(\omega_s)\rmd s\right\}\right]$. 
Following \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, if $\omega \sim\mathbb{W}_{\theta}^{k,X_{k}}$, $\kappa$ is a random variables taking values in $\mathbb{N}$ with distribution $\mu$ and if $(U_j)_{1\le j\le \kappa}$ are independent uniform random variables on $[0,t]$, 

@@

\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta t_k}\phi_{\theta}(\omega_s)\rmd s\right\}\right] &= \mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta t_k}\phi_{\theta}(\omega_s)\rmd s\right\}\right]\eqsp,\\
&= \mathrm{exp}\left\{-\Lambda_{\theta}\Delta t_k\right\}\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathrm{exp}\left\{\int_{0}^{\Delta t_k}(\Lambda_{\theta}-\phi_{\theta}(\omega_s))\rmd s\right\}\right]\eqsp,\\
&=\mathrm{exp}\left\{-\Lambda_{\theta}\Delta t_k\right\}\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\sum_{p\ge 0}\frac{1}{p!}\left(\int_{0}^{\Delta t_k}\Lambda_{\theta}-\phi_{\theta}(\omega_s)\rmd s\right)^p\right]\eqsp,\\
&=\mathrm{exp}\left\{-\Lambda_{\theta}\Delta t_k\right\}\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathbb{E}\left[\sum_{p\ge 0}\frac{(\Delta t_k)^p}{k!}\prod_{j=1}^p\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\right]\right]\eqsp,\\
&=\mathrm{exp}\left\{-\Lambda_{\theta}\Delta t_k\right\}\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\frac{(\Delta t_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\right]\eqsp.

@@

\Lambda_k = \left\{\kappa,\omega,U_1,\ldots,U_\kappa\right\}\eqsp,

@@

Algorithm~\ref{alg:AR} is used with 
\widehat{q}^{\Delta t_k}_{\theta}(X_{k-1},X_k;\Lambda) = \widetilde{q}^{\Delta t_k}_{\theta}(X_{k-1},X_k) \eqsp\mathrm{exp}\left\{H_{\theta}(X_k) - H_{\theta}(X_{k-1})-\ell(\theta)\Delta t_k\right\}\\ 
\times\mathrm{exp}\left\{-\Lambda_{\theta}\Delta t_k\right\}\frac{(\Delta t_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\eqsp.

>>>>

\newcommand{\x}[2]{x_{#1}^{(#2)}}
\newcommand{\w}[2]{w_{#1}^{(#2)}}
\newcommand{\tw}[2]{\tilde{w}_{#1}^{(#2)}}
\newcommand{\p}[2]{\xi_{#1}^{(#2)}}
\newcommand{\tp}[2]{\tilde{\xi}_{#1}^{(#2)}}
\newcommand{\ta}[2]{\tau_{#1}^{(#2)}}
\newcommand{\rmd}{\mathrm{d}}
\newcommand{\eqsp}{\;}
\newcommand{\1}{\mathrm{1}}
\newcommand{\com}[1]{{\color{gray} // #1}}
\newcommand{\mP}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\qk}{q_{k}}
\newcommand{\acom}[1]{\textit{\color{gray} //#1}}
\newcommand{\Oz}{Z}%Letter for the Ozaki approximation
\newcommand{\Jk}{J_{\alpha_\theta}^k}%Command for Jacobian of alpha
\newcommand{\mw}{\mathsf{w}}%For bridge realisations
%\newcommand{\qk}{q^{\Delta t_k}_{\theta}}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}

@@

For all $1\le k \le n$, the distribution of $X_{k+1} $ conditional on $X_{k}$ has a density $\qk(X_{k},\cdot) :=q(\theta,\Delta_k,X_{k-1},\cdot),~\Delta_k:=t_{k+1}-t_{k}$ with respect to a reference measure $\mu$ on $\mathbb{R}^d$.\\

@@

p_{\theta}(x_{0:n},Y_{0:n}) = \chi(x_0)g_{0}(x_0)\prod^{n-1}_{k=0}\qk(x_{k},x_{k+1})g_{k+1}(x_{k+1})\eqsp.

@@

+ \sum_{k=0}^{n-1}\mathbb{E}_{\theta_p}\left[\log \left(\qk(X_{k},X_{k+1})g_{k+1}(X_{k+1})\right)\middle|Y_{0:n}\right] \eqsp.

@@

H_{n} =  \sum_{k=0}^{n-1}\mathbb{E}\left[h_k(X_{k},X_{k+1})\middle|Y_{0:n}\right]\eqsp,

@@

when the transition density of the hidden states is not available explicitly and where $\{h_k\}_{k=0}^{n-1}$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$. The algorithm is based on the following link between the filtering and smoothing distributions:
\phi_{0:n|n}[h] = \phi_n[\E\left[h(X_{0:n})\vert X_n,~Y_{0:n}]\right]:= \phi_n[T_n(h)] \eqsp.\label{eq:FFbsm:equality}
Therefore, using the additive property and the usual filtering recursion applied on the right hand side of equation \eqref{eq:FFbsm:equality}, an approximation of \eqref{def:addfunc} can be computed using Sequential Monte Carlo techniques.\\ 

@@

The PaRIS algorithm uses the same decomposition as the FFBS algorithm introduced in \cite{doucetgodsillandrieu:2000} and the FFBSi algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions. 

@@

The focus is based sufficient statistics $\tau^i_k$, approximations of $T_k(h)$, for $1\le i\le N$ and $0\le k \le n$ starting with $\tau^i_0 = 0$ for all $1\le i\le N$. Let $\tilde{N}\ge 1$. Then, at each time step $0\le k \le {n-1}$ these statistics are updated according to the following steps.

@@

\item \label{it:PaRIS:sampleindex} For all $1\le i \le N$, sample independently $J_{k}^{i,\ell}$ in $\{1,\ldots,N\}$ for $1\le \ell \le \widetilde N$ with probabilities $\Lambda_{k}^N(i,\cdot)$, given by

@@

\Lambda_{k}^N(i,\ell) = \frac{\omega^{\ell}_{k} \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})}{\sum_{\ell=1}^N\omega^{\ell}_{k} \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})}\eqsp,\quad 1\le \ell\le N\eqsp.

@@

\tau^{i}_{k+1} := \frac{1}{\widetilde{N}} \sum^{\widetilde{N}}_{\ell=1} \left\{ \tau^{J_{k}^{i,\ell}}_{k} + h_{k} \left(\xi^{J_{k}^{i,\ell}}_{k}, \xi^{i}_{k+1}\right)  \right\}\eqsp.

@@

As proved in \cite{olsson:westerborn:2016}, the algorithm is asymptotically consistent (as $N$ goes to infinity) for any fixed precision parameter $\tilde N$. However, there is a significant qualitative difference between the cases $\tilde{N} = 1$ and $\tilde{N} \geq 2$.\\
As for the FFBSi algorithm,  when there exists $\sigma_+$ such that $0<\qk <\sigma_+$, the PaRIS algorithm may be implemented with $\mathcal{O}(N)$ complexity using the procedure of \cite{douc:garivier:moulines:olsson:2011} recalled in algorithm \ref{alg:AR:knownq}.
\begin{algorithm}[hb]
\caption{Random weight Accept-reject-based backward sampling  (\cite{douc:garivier:moulines:olsson:2011})}
\FORALL{$\ell \in 1,\dots, \widetilde N$} 
\STATE \textbf{\sc Sampling Step} Sample independently\\ $j\sim \{\omega_{k}^1,\dots,\omega_{k}^N\}$, $u\sim \mathcal{U}[0,1]$
\IF{ $$u \leq \frac{\qk(\xi_{k}^j,\xi_{k+1}^i,z)}{\sigma_+},$$}
\STATE Set $J_k^{i,\ell} = j$.
\STATE Return to \textbf{\sc Sampling Step}
\label{alg:AR:knownq}

%\begin{algorithm}
%\caption{Random weight Accept-reject-based backward sampling  (\cite{douc:garivier:moulines:olsson:2011})}
%\begin{algorithmic}
%\FORALL{$i \in 1,\dots,\widetilde N$}
%\STATE Set $L =\{1,\ldots,N\}$. 
%\WHILE{$L\neq\emptyset$}
%\STATE $p = \#L$.
%\STATE Sample $I = (I_1,\ldots,I_p)$ independently in $\{1,\ldots,N\}$ with probabilities proportional to $\omega_{k-1}^{\ell}$, $1\le \ell\le N$.
%\STATE Sample $U = (U_1,\ldots,U_p)$ independently and independently of $I$ uniformly in $[0,1]$.
%\STATE Sample $\Lambda = (\Lambda_1,\ldots,\Lambda_p)$ independently and independently of $I$ and $U$.
%\STATE Set $L_p = \emptyset$.
%\FORALL{$\ell \in 1,\dots,p$}
%\IF{$U_\ell<\widehat{q}(\xi_{k-1}^{I_\ell},\xi_k^{L(\ell)};\Lambda_{\ell})/\sigma_+$}
%\STATE $J_k^{L(\ell),i} = I_\ell$.
%\ELSE
%\STATE $L_p = L_p \cup \{L(\ell)\}$.
%\ENDIF
%\ENDFOR
%\STATE $L = L_p$.
%\ENDWHILE
%\ENDFOR
%\end{algorithmic}
%\label{alg:AR}
%\end{algorithm}
\section{The OzIRis algorithm}
\label{sec:PaRIS:SDE}
In general situations, the PaRIS algorithm cannot be used for stochastic differential equations as $\qk$ is unknown. Therefore, the computation of the importance weights $\omega_{k}^{\ell}$ and of the acceptance ratio in Algorithm~\ref{alg:AR:knownq} cannot be performed to apply the PaRIS algorithm. Following \cite{olsson:strojby:2011},
an extension of the PaRIS algorithm may be developed by replacing $\qk$  by an  estimator $\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)$, where $\zeta_k$ is a random variable such that:

@@

\mathbb{E}\left[\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta)\middle| \mathcal{F}_{k+1}^N\right] = \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})\eqsp,

@@

\mathcal{F}_{k+1}^N = \sigma\left\{Y_{0:k+1};(\xi^{\ell}_{u},\omega^{\ell}_{u},\tau^{\ell}_{u}),~1\le \ell\le N,~0\le u\le k+1\right\}\eqsp.

@@

In the case where $\widehat{q}_k$ may be upper bounded by a constant $\hat{\sigma}_+$, step~\eqref{it:PaRIS:sampleindex} is replaced by the acceptance-rejection mechanism given in Algorithm~\ref{alg:AR:unknownq}. The Lemma~\ref{lem:AR:unbiased} states that replacing  $\qk$ by its unbiased approximation is a valid alternative to step~\eqref{it:PaRIS:sampleindex}.
\begin{lemma}
\label{lem:AR:unbiased}
Let $0\le k\le n-1$.  For all $1\le i \le N$ and all $1\le \ell \le \widetilde{N}$, the conditional probability distribution given $\mathcal{F}_{k+1}^N$ of the random variable $J_{k}^{i,\ell}$ produced by Algorithm~\ref{alg:AR:unknownq} is  $\Lambda_{k}^N(i,\cdot)$. 
\end{lemma}
\begin{proof}
See Appendix \ref{sec:append:proofs}.
\begin{algorithm}[hb]
\caption{Double random weight Accept-reject-based backward sampling}
\FORALL{$\ell \in 1,\dots, \widetilde N$} 
\STATE \textbf{\sc Sampling Step} Sample independently\\ $j\sim \{\omega_{k}^1,\dots,\omega_{k}^N\}$, $z\sim\zeta_k$, $u\sim \mathcal{U}[0,1]$
\IF{ $$u \leq \frac{\widehat{\qk}(\xi_{k}^j,\xi_{k+1}^i,z)}{\hat{\sigma}_+},$$}
\STATE Set $J_k^{i,\ell} = j$.
\STATE Return to \textbf{\sc Sampling Step}
\label{alg:AR:unknownq}
In \cite{olsson:strojby:2011},  Global Poisson Estimator (GPE) are used to estimate $\qk$ which relies on a representation of $\qk$ as an expectation under the probability measure of a Brownian bridge. 
In order to improve the Monte Carlo estimation of $\qk$,  other diffusion bridges may be used as long as they can be simulated efficiently.\\ 
A simple idea is to consider a diffusion obtained as an approximation of the target SDE \eqref{eq:target:sde}. 
\cite{ozaki:1992,shoji:ozaki:1998} proposed a higher order approximation of the drift term $\alpha_{\theta}$ than the Euler scheme. 
The authors used a linear approximation of the diffusion drift $\alpha_{\theta}$, together with a constant approximation of the volatility over each time interval $(t_{k},t_{k+1})$.\\ 
Formally, for all $0\le k\le n-1$, let $(Z^{k}_t)_{t\ge 0}$ be governed by the following stochastic differential equation for $t\in(t_{k},t_{k+1})$:
\label{eq:SDE:Ozaki}
\Oz^k_{t_{k}} = X_{k}\quad\mbox{and}\quad\rmd \Oz^k_t = \widetilde{\alpha}^k_{\theta}(\Oz^k_t)\rmd t + \rmd W_t\eqsp,
\widetilde{\alpha}^k_{\theta}(x) = \Jk x + \alpha_{\theta}(X_{k}) - \Jk X_{k}
and $\Jk$ is the Jacobian of $\alpha_{\theta}$ evaluated in $X_k$. The probability density function of the transition kernel associated with $\Oz^k$ is written $\widetilde{q}_k$ and is given in Section~\ref{sec:ornstein:bridge}.
Writing $\mathbb{Q}^k_{\theta}$ (resp. $\mathbb{W}^{k}_{\theta}$) the probability measure induced by $(X_s)_{t_{k}\le s\le t_{k+1}}$  (resp. $(\Oz_s)_{t_{k}\le s\le t_{k+1}}$) on $(\mathsf{C},\mathcal{C})$, Girsanov theorem yields:
\frac{\rmd \mathbb{Q}^k_{\theta}}{\rmd \mathbb{W}^{k}_{\theta}}(w) = \mathrm{exp}\left\{H_{\theta}(\omega_{t_{k+1}}) - H_{\theta}(\omega_{t_{k}}) - \frac{1}{2}\int_{t_{k}}^{t_{k+1}}\left\{\|\alpha_{\theta}(\mw_s)-\widetilde{\alpha}_{\theta}(\mw_s)\|^2+\Delta H_{\theta}(\mw_s)\right\}\rmd s\right\}\eqsp,
H_{\theta}(x) = A_{\theta}(x) - \frac{1}{2}x'\Jk x- (\alpha_{\theta}(X_{k}) - \Jk X_{k-1})'x
and $\Delta$ is the Laplace operator. Then, the probability measures conditioned on hitting $X_{k+1}$ at time $t_{k+1}$ satisfy:

@@

\frac{\rmd \mathbb{Q}^{k,X_{k+1}}_{\theta}}{\rmd \mathbb{W}^{k,X_{k+1}}_{\theta}}(\mw) = \frac{\widetilde{\qk}(X_{k},X_{k+1})}{\qk (X_{k},X_{k+1}) }\mathrm{exp}\left\{H_{\theta}(X_{k+1}) - H_{\theta}(X_{k})\right\} \\
\times \mathrm{exp}\left\{- \frac{1}{2}\int_{t_{k}}^{t_{k+1}}\left\{\|\alpha_{\theta}(\mw_s)-\widetilde{\alpha}_{\theta}(\mw_s)\|^2+\Delta H_{\theta}(\mw_s)\right\}\rmd s\right\}\eqsp.

@@

\item For all $\theta\in\Theta$ and all $0\le k \le n-1$, $x\mapsto \alpha_{\theta}(x) - \alpha_{\theta}(X_k) - \Jk(x-X_k)$ is bounded.
\item For all $\theta\in\Theta$ and all $0\le k \le n-1$, $x\mapsto \mathrm{Tr}\left(J_{\alpha_{\theta}}(x) - \Jk\right)$ is bounded, where $\mathrm{Tr}$ is the trace operator.

@@

\qk(X_{k},X_{k+1}) = \widetilde{\qk}(X_{k},X_{k+1}) \eqsp\mathrm{exp}\left\{H_{\theta}(X_{k+1}) - H_{\theta}(X_{k})-\ell(\theta)\Delta_k\right\}\\
\times\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{ \Delta_k}\phi_{\theta}(\mw_s)\rmd s\right\}\right]\eqsp,

@@

where $\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}}$ is the expectation under the law of the bridge process defined by \eqref{eq:SDE:Ozaki}, starting at $X_{k}$ at time $0$ and ending at $X_{k+1}$ at time $\Delta_k$.
An unbiased estimator of $\qk(X_{k},X_{k+1})$ is then obtained by estimating $\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{ \Delta_k}\phi_{\theta}(\mw_s)\rmd s\right\}\right]$. 
Following \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, if $\omega \sim\mathbb{W}_{\theta}^{k,X_{k}}$, $\kappa$ is a random variable taking values in $\mathbb{N}$ with distribution $\mu$ and if $(U_j)_{1\le j\le \kappa}$ are independent uniform random variables on $[0,t]$, 

@@

\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta_k}\phi_{\theta}(\mw_s)\rmd s\right\}\right] &= \mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}} \left[\mathrm{exp}\left\{-\int_{0}^{\Delta_k}\phi_{\theta}(\mw_s)\rmd s\right\}\right]\eqsp,\\
&= \mathrm{exp}\left\{-\Lambda_{\theta} \Delta_k\right\} \mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}} \left[\mathrm{exp}\left\{ \int_{0}^{\Delta_k}(\Lambda_{\theta}-\phi_{\theta}(\mw_s))\rmd s\right\}\right]\eqsp,\\
&=\mathrm{exp}\left\{-\Lambda_{\theta}\Delta_k\right\} \mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}}\left[\sum_{p\ge 0}\frac{1}{p!}\left(\int_{0}^{\Delta_k}\Lambda_{\theta}-\phi_{\theta}(\mw_s)\rmd s\right)^p\right]\eqsp,\\
&=\mathrm{exp}\left\{-\Lambda_{\theta}\Delta_k \right\} \mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}} \left[\mathbb{E}\left[\sum_{p\ge 0}\frac{( \Delta_k)^p}{k!} \prod_{j=1}^p\left(\Lambda_{\theta}-\phi_{\theta}(\mw_{U_j})\right)\right]\right]\eqsp,\\
&=\mathrm{exp}\left\{-\Lambda_{\theta} \Delta_k\right\} \mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}}\left[ \frac{(\Delta_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\mw_{U_j})\right)\right]\eqsp.

@@

\zeta_k = \left\{\kappa,\mw,U_1,\ldots,U_\kappa\right\}\eqsp,

@@

Algorithm~\ref{alg:AR:unknownq} is used with 
\widehat{\qk}(X_{k},X_{k+1};\zeta_k) = \widetilde{\qk}(X_{k},X_{k+1}) \eqsp\mathrm{exp}\left\{H_{\theta}(X_{k+1}) - H_{\theta}(X_{k})-\ell(\theta) \Delta_k\right\}\\ 
\times\mathrm{exp}\left\{-\Lambda_{\theta}\Delta_k\right\}\frac{( \Delta_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\mw_{U_j})\right)\eqsp.\label{eq:unbiased:q}
The choice of $\mu$ is restricted to distributions such that the associated estimator is upper bounded by an explicit constant $\hat{\sigma}_+^k(\theta)$. Then, for all $0\le k \le n-1$, $1\le i\le N$, the importance weight $\omega_k^{\ell}$ is replaced by the unbiased estimator:
